---
title: "Environmental variables"
author: "Dana Lanceman"
date: "12/9/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Prepare data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/danal/OneDrive - UNSW/Desktop/WRL/R')
# replace the above with the path to your home directory
```

```{r get packages}
library(tidyverse)
library(raster)
library(sf)
library(padr)
library(Kendall)
library(lubridate)
library(ggpubr)
library(rstatix)
library(gdata)
library(zoo)
```


```{r get data}
# elevation data
dem <- raster("raw/20170213 kooragang island rgb_dsm.tif")

# cropped maps
zero1 <- raster("raw/0mos_classified_0_1_all.tif")
nine1 <- raster("raw/9mos_classified_evenmoremud.tif")
eighteen1 <- raster("raw/18mos_classified_01_all.tif")
twentysix1 <- raster("raw/26mos_classified_01_all.tif")
thirtyfour1 <- raster("raw/34mos_classified_01_all.tif")
fortytwo1 <- raster("raw/42mos_classified_01_all.tif")
fortysix1 <- raster("raw/46mos_classified_01_all.tif")

# 46 minus 0 months - sarc/spor loss and growth map
ch46minus0 <- raster("raw/change_tiffs/46mos_minus_0mos.tif")

# other loss and growth tiffs
ch46minus42t <- raster("raw/change_tiffs/46mos_minus_42mos.tif")
ch42minus34t <- raster("raw/change_tiffs/42mos_minus_34mos.tif")
ch34minus26t <- raster("raw/change_tiffs/34mos_minus_26mos.tif")
ch26minus18t <- raster("raw/change_tiffs/26mos_minus_18mos.tif")
ch18minus9t <- raster("raw/change_tiffs/18mos_minus_9mos_moremud.tif")
ch9minus0t <- raster("raw/change_tiffs/9mos_minus_0mos_moremud.tif")

# boundary for clipping 
boundary <- st_read('raw/Clip_work.shp')
boundary <- st_transform(boundary, CRS(proj4string(zero1)))
```


```{r prepare data}
# for 9 months, reclassify class 8 (sarc clumps) as class 2 (sarc)

# create 2 column reclassification matrix for integer data
matrix.data <- c(0,0,1,1,2,2,3,3,4,4,5,5,6,6,7,7,8,2)
matrix <- matrix(matrix.data, nrow = 9, ncol = 2, byrow = TRUE)
matrix

# use matrix for reclassifications
nine1 <- reclassify(nine1, matrix)



# clip all rasters to same extent
zero1c <- crop(zero1, extent(boundary)) 
zero1c <- mask(zero1c, boundary)

nine1c <- crop(nine1, extent(boundary)) 
nine1c <- mask(nine1c, boundary)

eighteen1c <- crop(eighteen1, extent(boundary)) 
eighteen1c <- mask(eighteen1c, boundary)

twentysix1c <- crop(twentysix1, extent(boundary)) 
twentysix1c <- mask(twentysix1c, boundary)

thirtyfour1c <- crop(thirtyfour1, extent(boundary)) 
thirtyfour1c <- mask(thirtyfour1c, boundary)

fortytwo1c <- crop(fortytwo1, extent(boundary)) 
fortytwo1c <- mask(fortytwo1c, boundary)

fortysix1c <- crop(fortysix1, extent(boundary)) 
fortysix1c <- mask(fortysix1c, boundary)



ch46minus0c <- crop(ch46minus0, extent(boundary)) 
ch46minus0c <- mask(ch46minus0c, boundary)

ch46minus42c <- crop(ch46minus42t, extent(boundary)) 
ch46minus42c <- mask(ch46minus42c, boundary)

ch42minus34c <- crop(ch42minus34t, extent(boundary)) 
ch42minus34c <- mask(ch42minus34c, boundary)

ch34minus26c <- crop(ch34minus26t, extent(boundary)) 
ch34minus26c <- mask(ch34minus26c, boundary)

ch26minus18c <- crop(ch26minus18t, extent(boundary)) 
ch26minus18c <- mask(ch26minus18c, boundary)

ch18minus9c <- crop(ch18minus9t, extent(boundary)) 
ch18minus9c <- mask(ch18minus9c, boundary)

ch9minus0c <- crop(ch9minus0t, extent(boundary)) 
ch9minus0c <- mask(ch9minus0c, boundary)




demc <- crop(dem, extent(boundary)) 
demc <- mask(demc, boundary)



# resample dem to same (coarser) resolution as classified images
# also resample all classified images to nine1c as they all have slightly different resolutions, and the same resolution is required for processing
dem_res <- raster::resample(demc, nine1c, method = "bilinear")
zero_res <- raster::resample(zero1c, nine1c, method = "ngb")
eighteen_res <- raster::resample(eighteen1c, nine1c, method = "ngb")
twentysix_res <- raster::resample(twentysix1c, nine1c, method = "ngb")
thirtyfour_res <- raster::resample(thirtyfour1c, nine1c, method = "ngb")
fortytwo_res <- raster::resample(fortytwo1c, nine1c, method = "ngb")
fortysix_res <- raster::resample(fortysix1c, nine1c, method = "ngb")

ch46minus0_res <- raster::resample(ch46minus0c, nine1c, method = "ngb")
ch46minus42_res <- raster::resample(ch46minus42c, nine1c, method = "ngb")
ch42minus34_res <- raster::resample(ch42minus34c, nine1c, method = "ngb")
ch34minus26_res <- raster::resample(ch34minus26c, nine1c, method = "ngb")
ch26minus18_res <- raster::resample(ch26minus18c, nine1c, method = "ngb")
ch18minus9_res <- raster::resample(ch18minus9c, nine1c, method = "ngb")
ch9minus0_res <- raster::resample(ch9minus0c, nine1c, method = "ngb")

```


```{r visualise}
plot(dem_res)

plot(zero_res)
plot(nine1c)
plot(eighteen_res)
plot(twentysix_res)
plot(thirtyfour_res)
plot(fortytwo_res)
plot(fortysix_res)
```

# Elevation - calculations / data manipulation

Calculate zonal statistics 

I'm just using means, but SD, min, max, sum and count can also be calculated

```{r zero months}
zero_mean <- zonal(dem_res, zero_res, fun = "mean", digits = 2) %>% 
  as.data.frame() %>% 
  mutate(dataset = 0,
         category = c("na", "spor","sarc", "grass", "mud", "water", "white_vege","road"))
```

```{r nine months}
nine_mean <- zonal(dem_res, nine1c, fun = "mean", digits = 2) %>% 
  as.data.frame() %>% 
  mutate(dataset = 9,
         category = c("na", "spor", "sarc",  "grass", "mud", "water", "white_vege", "road"))
```

```{r eighteen months}
eighteen_mean <- zonal(dem_res, eighteen_res, fun = "mean", digits = 2) %>% 
  as.data.frame() %>% 
  mutate(dataset = 18,
         category = c("na", "spor","sarc", "grass", "mud", "water","sueda","road",  "white_vege")) 
```

```{r twenty six months}
twentysix_mean <- zonal(dem_res, twentysix_res, fun = "mean", digits = 2) %>%
  as.data.frame() %>%  
  mutate(dataset = 26,
         category = c("spor","sarc", "grass", "mud", "water","sueda", "road", "white_vege"))
```

```{r thirty four months}
thirtyfour_mean <- zonal(dem_res, thirtyfour_res, fun = "mean", digits = 2) %>% 
  as.data.frame() %>% 
  mutate(dataset = 34,
         category = c("spor","sarc","grass", "mud", "water","sueda","road", "white_vege"))
```

```{r forty two months}
fortytwo_mean <- zonal(dem_res, fortytwo_res, fun = "mean", digits = 2) %>% 
  as.data.frame() %>% 
  mutate(dataset = 42,
         category = c("na", "spor","sarc","grass", "mud", "water","algae","road", "white_vege", "sueda"))
```

```{r forty six months}
fortysix_mean <- zonal(dem_res, fortysix_res, fun = "mean", digits = 2) %>% 
  as.data.frame() %>% 
  mutate(dataset = 46,
         category = c("spor","sarc","grass", "mud", "water","sueda","road", "white_vege"))
```

```{r combine, make wide format}
allzonal <- rbind(zero_mean, nine_mean, eighteen_mean, twentysix_mean, thirtyfour_mean, fortytwo_mean, fortysix_mean) %>% 
  filter(category %in% c("spor", "sarc", "sueda")) %>% 
  select(!zone)
```

```{r add date column}
allzonal <- allzonal %>% mutate(date = case_when(
         dataset == 0 ~ "2017/02/03",
         dataset == 9 ~ "2017/10/19",
         dataset == 18 ~ "2018/08/22",
         dataset == 26 ~ "2019/04/30",
         dataset == 34 ~ "2019/12/18",
         dataset == 42 ~ "2020/08/13",
         dataset == 46 ~ "2020/12/08")) 
allzonal$date <- as.Date(allzonal$date)
```


# Elevation - plot

does the elevation that different saltmarsh grows at differ between species and over time?

```{r plot function}
plot_elev <- function(variable){
  ggplot(allzonal) +
  aes(x = date, y = variable, color = category, linetype = category) +
  geom_line(size = 3) +
    scale_color_manual(values = c("#1fbfad", "#f2c80c", "#5c0b1f"), labels = c("Sarcocornia","Sporobolus","Suaeda")) +
    scale_linetype_manual(values=c(2,3,4), labels = c("Sarcocornia","Sporobolus","Suaeda")) +
    labs(x = "Date", y = "Mean elevation (m)", color = "", linetype = "") +
  theme_classic() +
  guides(color = guide_legend(byrow = TRUE)) + # for legend spacing 
  theme(axis.text = element_text(size = 40),
        axis.title = element_text(size = 54),
        axis.text.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.text.x = element_text(vjust = -1),
        axis.title.x = element_text(vjust = -1),
        legend.text = element_text(size = 30),
        legend.title = element_text(size = 36),
        plot.margin = margin(20,20,20,10),
        legend.spacing.y = unit(1, "cm"),
        legend.key.size = unit(3,"line"))
}
```


```{r plot mean elevation}
plot_elev(allzonal$mean)
# export at 2100 x 1400
```



# Elevation Trend analysis 
Did mean elevation change over time for each species?

```{r pad time series}
allzonal.spor <- allzonal %>% filter(category == "spor") %>% pad()
allzonal.sarc <- allzonal %>% filter(category == "sarc") %>% pad() # sig
allzonal.sueda <- allzonal %>% filter(category == "sueda") %>% pad()
```

```{r Kendall rank correlations}
MannKendall(ts(allzonal.spor$mean,frequency = 365, start = c(2017,34)))
MannKendall(ts(allzonal.sarc$mean,frequency = 365, start = c(2017,34)))
MannKendall(ts(allzonal.sueda$mean,frequency = 365, start = c(2018,313)))

```




# Elevation - compare elevation of areas of saltmarsh growth vs loss

Calculate mean and SD in this scenario, to visualise the distribution of elevations that different processes are occuring at

```{r 46 mos minus 0 mos}
ch_mean <- zonal(dem_res, ch46minus0_res, fun = "mean", digits = 2) 
ch_sd <- zonal(dem_res, ch46minus0_res, fun = "sd", digits = 2)

ch_demstats <- combine(ch_mean, ch_sd) 
ch_demstats$source <-  recode(ch_demstats$source, 
         "ch_mean" = "mean",
         "ch_sd" = "sd") 
 
ch_demstats$category <- c("Sarcocornia loss", "Sarcocornia to Sporobolus","Sporobolus loss", "No change", "Sporobolus growth", "Sporobolus to Sarcocornia", "Sarcocornia growth")
```

```{r make into wide format}
ch_demstats_wide <- ch_demstats %>% dplyr::select(mean, source, category) %>%
  pivot_wider(names_from = source, values_from = mean) %>% 
  filter(category != "No change")
```


```{r plot}
# graph with SD range
ggplot(ch_demstats_wide, aes(category, mean, ymin = mean-sd, ymax = mean+sd)) +
  geom_point(size = 5) +
  geom_errorbar(size = 2, width = 0.5) +
  labs(y = "Mean elevation (m)", x = "") +
  theme_bw() +
  scale_x_discrete(labels= c("Sarc growth", "Sarc loss", "Sarc to Spor", "Spor growth", "Spor loss", "Spor to Sarc")) +
  theme(axis.title = element_text(size = 48),
        axis.text = element_text(size = 36),
        axis.text.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.text.x = element_text(vjust = -1),
        panel.grid.major = element_blank(), # remove grid lines
        panel.grid.minor = element_blank())
```


# Sarc and spor growth/loss over diff time periods vs elevation


```{r 46 mos minus 42 mos}
ch42_mean <- zonal(dem_res, ch46minus42_res, fun = "mean", digits = 2) %>% 
  as.data.frame() %>% 
  mutate(dataset = 46,
         category = c("Sarcocornia loss", "Sarcocornia to Sporobolus","Sporobolus loss", "No change", "Sporobolus growth", "Sporobolus to Sarcocornia", "Sarcocornia growth"))
```


```{r 42 mos minus 34 mos}
ch34_mean <- zonal(dem_res, ch42minus34_res, fun = "mean", digits = 2) %>%  
  as.data.frame() %>% 
  mutate(dataset = 42, 
         category = c("Sarcocornia loss", "Sarcocornia to Sporobolus","Sporobolus loss", "No change", "Sporobolus growth", "Sporobolus to Sarcocornia", "Sarcocornia growth"))
```


```{r 34 mos minus 26 mos}
ch26_mean <- zonal(dem_res, ch34minus26_res, fun = "mean", digits = 2) %>%  
  as.data.frame() %>% 
  mutate(dataset = 34, 
         category = c("Sarcocornia loss", "Sarcocornia to Sporobolus","Sporobolus loss", "No change", "Sporobolus growth", "Sporobolus to Sarcocornia", "Sarcocornia growth"))
```


```{r 26 mos minus 18 mos}
ch18_mean <- zonal(dem_res, ch26minus18_res, fun = "mean", digits = 2) %>%  
  as.data.frame() %>% 
  mutate(dataset = 26, 
         category = c("Sarcocornia loss", "Sarcocornia to Sporobolus","Sporobolus loss", "No change", "Sporobolus growth", "Sporobolus to Sarcocornia", "Sarcocornia growth"))
```


```{r 18 mos minus 9 mos}
ch18_9_mean <- zonal(dem_res, ch18minus9_res, fun = "mean", digits = 2) %>%  
  as.data.frame() %>% 
  mutate(dataset = 18,
         category = c("Sarcocornia loss", "Sarcocornia to Sporobolus","Sporobolus loss", "No change", "Sporobolus growth", "Sporobolus to Sarcocornia", "Sarcocornia growth"))
```


```{r 9 mos minus 0 mos}
ch9_0_mean <- zonal(dem_res, ch9minus0_res, fun = "mean", digits = 2) %>%  
  as.data.frame() %>% 
  mutate(dataset = 9, 
         category = c("Sarcocornia loss", "Sarcocornia to Sporobolus","Sporobolus loss", "No change", "Sporobolus growth", "Sporobolus to Sarcocornia", "Sarcocornia growth"))
```



```{r combine change datasets}
growthzonal <- rbind(ch42_mean, ch34_mean, ch26_mean, ch18_mean, ch18_9_mean, ch9_0_mean) %>% 
  filter(category != "No change") %>% 
  select(!zone)
```

```{r add date column}
growthzonal <- growthzonal %>% mutate(date = case_when(
         dataset == 9 ~ "2017/10/19",
         dataset == 18 ~ "2018/08/22",
         dataset == 26 ~ "2019/04/30",
         dataset == 34 ~ "2019/12/18",
         dataset == 42 ~ "2020/08/13",
         dataset == 46 ~ "2020/12/08")) 
growthzonal$date <- as.Date(growthzonal$date)
```


```{r plot}
ggplot(growthzonal) +
  aes(x = dataset, y = mean, color = category, linetype = category) +
  geom_line(size = 3) +
    labs(x = "Months since restoration", y = "Mean elevation (m)", color = "", linetype = "") +
  #scale_x_date(date_breaks = "1 year", date_labels =  "%Y") +
  theme_classic() +
  guides(color = guide_legend(byrow = TRUE)) + # for legend spacing 
  theme(axis.text = element_text(size = 40),
        axis.title = element_text(size = 54),
        legend.text = element_text(size = 30),
        legend.title = element_text(size = 36),
        plot.margin = margin(20,20,20,10),
        legend.spacing.y = unit(1, "cm"),
        legend.key.size = unit(3,"line"),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)))

# export at 2100 x 1400
```


# Memory issues?

```{r garbage collection}
gc() # garbage collection to help free up some memory
```

