---
title: "patch_metrics"
author: "Dana Lanceman"
date: "12/8/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/danal/OneDrive - UNSW/Desktop/WRL/R')
```

```{r get packages}
library(tidyverse)
library(raster)
library(sf)
library(landscapemetrics)
library(viridis)
library(ggpubr)
library(Kendall)
library(padr)

# note that landscapemetrics requires a lot of memory and may come up with an "rcpp" package error if the hardware is not sufficient. Try with a computer with more memory in that case! 

# I've run this and output the patch metrics as two csvs - see chunk 10. Can just skip to this chunk to save time or if insufficient memory. 
```


```{r load data - just 0.1 resolution}
zero1 <- raster("raw/0mos_classified_0_1_all.tif")
nine1 <- raster("raw/9mos_classified_01_all_new.tif")
eighteen1 <- raster("raw/18mos_classified_01_all.tif")
twentysix1 <- raster("raw/26mos_classified_01_all.tif")
thirtyfour1 <- raster("raw/34mos_classified_01_all.tif")
fortytwo1 <- raster("raw/42mos_classified_01_all.tif")
fortysix1 <- raster("raw/46mos_classified_01_all.tif")

# get shapefile and convert to same coordinate system
# need to replace the #'s with a real file name
boundary <- st_read('raw/Clip_work.shp')
boundary <- st_transform(boundary, CRS(proj4string(zero1)))
```

```{r visualise data}
plot(zero1)
plot(nine1)
plot(eighteen1)
plot(twentysix1) # background is labelled # 5 - need to remove
plot(thirtyfour1)
plot(fortytwo1)
plot(fortysix1) # background is labelled # 5 - need to remove

plot(boundary)
```

```{r crop out background for 26 mos and 46 mos}
# crop 26 mos
twsix1.crop <- crop(twentysix1, extent(boundary)) 
twsix1.crop <- mask(twsix1.crop, boundary)
plot(twsix1.crop)

# crop 46 mos
fosix1.crop <- crop(fortysix1, extent(boundary)) 
fosix1.crop <- mask(fosix1.crop, boundary)
plot(fosix1.crop)
```

```{r reassign sarc_clumps as sarc in 9 mos}
# reclassify class 8 as class 2

# create 2 column reclassification matrix for integer data
matrix.data <- c(0,0,1,1,2,2,3,3,4,4,5,5,6,6,7,7,8,2)
matrix <- matrix(matrix.data, nrow = 9, ncol = 2, byrow = TRUE)
matrix

# use matrix for reclassifications
nine1 <- reclassify(nine1, matrix)

# visualise
plot(nine1)
```


Landscape metrics package
- article explaining: https://onlinelibrary.wiley.com/doi/full/10.1111/ecog.04617
- documentation: https://cran.r-project.org/web/packages/landscapemetrics/landscapemetrics.pdf
```{r check data are suitable for landscapemetrics}
check_landscape(zero1)
check_landscape(nine1)
check_landscape(eighteen1)
check_landscape(twsix1.crop)
check_landscape(thirtyfour1)
check_landscape(fortytwo1)
check_landscape(fosix1.crop)
# all ok
```


```{r prepare for metric calculation}
possible_metrics <- list_lsm() # list of all possible metrics
#view(possible_metrics)

# make a list of rasters
raster_names <- list(zero1, nine1, eighteen1, twsix1.crop, thirtyfour1, fortytwo1, fosix1.crop)

# make a function to add dataset names and categsories to data
add.metadata <- function(x){
  x %>% mutate(
  dataset = case_when(id == 1 ~ "zero",
                                id == 2 ~ "nine",
                                id == 3 ~ "eighteen",
                                id == 4 ~ "twentysix",
                                id == 5 ~ "thirtyfour",
                                id == 6 ~ "fortytwo",
                                id == 7 ~ "fortysix"),
  category = c("na", "spor","sarc", "grass", "mud", "water", "white_vege","road", 
               "na", "spor", "sarc", "grass", "mud", "water", "white_vege","road", 
               "na", "spor","sarc", "grass", "mud", "water","sueda","road",  "white_vege",
               "spor","sarc", "grass", "mud", "water","sueda", "road","white_vege",
               "na", "spor","sarc","grass", "mud", "water","sueda","road", "white_vege",
               "na", "spor","sarc","grass", "mud", "water","algae","road", "white_vege", "sueda",
               "spor","sarc","grass", "mud", "water","sueda","road", "white_vege")) %>% 
  filter(category != "na") # remove na data
}
# no na for 26 mos / 46 mos due to cropping


# function for adding metadata for landscape metrics
add.meta.land <- function(x){
  x %>% mutate(
  dataset = case_when(id == 1 ~ "zero",
                                id == 2 ~ "nine",
                                id == 3 ~ "eighteen",
                                id == 4 ~ "twentysix",
                                id == 5 ~ "thirtyfour",
                                id == 6 ~ "fortytwo",
                                id == 7 ~ "fortysix"))
}
```

```{r calculate patch metrics}
# apply functions to all datasets at once - note that each takes minutes

# CLASS METRICS
# area and edge metrics
area.means <- lapply(raster_names, lsm_c_area_mn) %>% bind_rows(.id = "id") %>% add.metadata()  # patch area, mean
area.sds <- lapply(raster_names, lsm_c_area_sd) %>% bind_rows(.id = "id") %>% add.metadata() # patch area, sd
area.totals <- lapply(raster_names, lsm_c_ca) %>% bind_rows(.id = "id") %>% add.metadata() # total class area

# core area metrics
cai.mns <- lapply(raster_names, lsm_c_cai_mn) %>% bind_rows(.id = "id") %>% add.metadata() # core area index, mean
cai.sds <- lapply(raster_names, lsm_c_cai_sd) %>% bind_rows(.id = "id") %>% add.metadata() # core area index, sd
core.mns <- lapply(raster_names, lsm_c_core_mn) %>% bind_rows(.id = "id") %>% add.metadata() # alternative core area, mean
core.sds <- lapply(raster_names, lsm_c_core_sd) %>% bind_rows(.id = "id") %>% add.metadata() # alternative core area, sd

# aggregation metrics
nps <- lapply(raster_names, lsm_c_np) %>% bind_rows(.id = "id") %>% add.metadata() # number of patches
cohesions <- lapply(raster_names, lsm_c_cohesion) %>% bind_rows(.id = "id") %>% add.metadata() # cohesion
enn.mns <- lapply(raster_names, lsm_c_enn_mn) %>% bind_rows(.id = "id") %>% add.metadata() # euclidean nearest neighbour distance, mean
enn.sds <- lapply(raster_names, lsm_c_enn_sd) %>% bind_rows(.id = "id") %>% add.metadata() # euclidean nearest neighbour distance, sd
clumpys <- lapply(raster_names, lsm_c_clumpy) %>% bind_rows(.id = "id") %>% add.metadata() # clumpiness

# shape metrics
contig.mns <- lapply(raster_names, lsm_c_contig_mn) %>% bind_rows(.id = "id") %>% add.metadata() # contiguity, mean
contig.sds <- lapply(raster_names, lsm_c_contig_sd) %>% bind_rows(.id = "id") %>% add.metadata() # contiguity, sd
shape.mns <- lapply(raster_names, lsm_c_shape_mn) %>% bind_rows(.id = "id") %>% add.metadata() # shape index, mean
shape.sds <- lapply(raster_names, lsm_c_shape_sd) %>% bind_rows(.id = "id") %>% add.metadata() # shape index, sd


# need to put landscape metrics in a diff table- not compatible - so recode

# LANDSCAPE METRICS
# aggregation metrics
contags <- lapply(raster_names, lsm_l_contag) %>% bind_rows(.id = "id") %>% add.meta.land() # contagion

# diversity metrics
shan.divs <- lapply(raster_names, lsm_l_shdi) %>% bind_rows(.id = "id") %>% add.meta.land() # shannon diversity
shan.evens <- lapply(raster_names, lsm_l_shei) %>% bind_rows(.id = "id") %>% add.meta.land() # shannon evenness 






# combine outputs   
allpatchc <- rbind(area.means, area.sds, area.totals, core.mns, core.sds, cai.mns, cai.sds, nps, cohesions, enn.mns, enn.sds, clumpys, contig.mns, contig.sds, shape.mns, shape.sds) # bind all class metric outputs into one tibble

allpatchl <- rbind(contags, shan.divs, shan.evens)

# then change to wide format with each patch metric as its own column
# add numerical column for dataset (months since restoration)
allpatch.w <- allpatchc %>% dplyr::select(metric, value, dataset, category) %>% pivot_wider(names_from = metric, values_from = value) %>% 
  mutate(months = case_when(dataset == "zero" ~ 0,
                            dataset == "nine" ~ 9,
                            dataset == "eighteen" ~ 21,
                            dataset == "twentysix" ~ 26,
                            dataset == "thirtyfour" ~ 34,
                            dataset == "fortytwo" ~ 42,
                            dataset == "fortysix" ~ 46))

allpatchl.w <- allpatchl %>% dplyr::select(metric, value, dataset) %>% pivot_wider(names_from = metric, values_from = value) %>% 
  mutate(months = case_when(dataset == "zero" ~ 0,
                            dataset == "nine" ~ 9,
                            dataset == "eighteen" ~ 21,
                            dataset == "twentysix" ~ 26,
                            dataset == "thirtyfour" ~ 34,
                            dataset == "fortytwo" ~ 42,
                            dataset == "fortysix" ~ 46))
```


```{r import patch metrics data}
# generated from above script. Just because it takes so long to generate and doesn't work on computers with insufficient memory

allpatch.w <- read.csv("export/csv/classmetrics.csv")
allpatchl.w <- read.csv("export/csv/landmetrics.csv")

# create a date column
allpatch.w <- allpatch.w %>% mutate(date = case_when(
         months == 0 ~ "2017/02/03",
         months == 9 ~ "2017/10/19",
         months == 21 ~ "2018/11/09",
         months == 26 ~ "2019/04/30",
         months == 34 ~ "2019/12/18",
         months == 42 ~ "2020/08/13",
         months == 46 ~ "2020/12/08")) 
allpatch.w$date <- as.Date(allpatch.w$date)

allpatchl.w <- allpatchl.w %>% mutate(date = case_when(
         months == 0 ~ "2017/02/03",
         months == 9 ~ "2017/10/19",
         months == 21 ~ "2018/11/09",
         months == 26 ~ "2019/04/30",
         months == 34 ~ "2019/12/18",
         months == 42 ~ "2020/08/13",
         months == 46 ~ "2020/12/08")) 
allpatchl.w$date <- as.Date(allpatchl.w$date)
```


# visualisation

probably just visualise saltmarsh classes! could have made an overall saltmarsh class first... but didn't. can consider later if we really want it and want this to run for another 3 1/2 hours

```{r subset for only saltmarsh classes}
allpatch.ws <- allpatch.w %>% filter(category %in% c("spor", "sarc", "sueda")) %>% mutate(category = as.factor(category))

# can't do this for landscape metrics, as it's  about the entire landscape with all classes included
```

```{r function for plotting}
patch_plot <- function(y_var, ylab){
  allpatch.ws %>% 
  ggplot() +
    aes(x = date, y = y_var, color = category, linetype = category) +
  geom_line(size = 3) +
    scale_color_manual(values = c("#1fbfad", "#f2c80c", "#5c0b1f"), labels = c("Sporobolus virginicus","Sarcocornia quinqueflora","Suaeda australis")) +
    scale_linetype_manual(values=c(2,3,4), labels = c("Sporobolus virginicus","Sarcocornia quinqueflora","Suaeda australis")) +
    labs(x = "Date", y = ylab, color = "", linetype = "") +
  theme_classic() +
  theme(text = element_text(size = 36),
        axis.text = element_text(size = 36),
        axis.title = element_text(size = 40),
        legend.text = element_text(size = 36, face = "italic"),
        plot.margin = margin(20,20,10,10),
        legend.key.size = unit(6,"line"),
        axis.text.x = element_text(vjust = -1),
        axis.title.x = element_text(vjust = -1))
}

land_plot <- function(y_var, ylab){
  allpatchl.w %>% 
  ggplot() +
    aes(x = date, y = y_var) +
  geom_line(size = 3) +
    labs(x = "Date", y = ylab) +
  theme_classic() +
  theme(axis.text = element_text(size = 40),
        axis.title = element_text(size = 48),
        plot.margin = margin(20,20,20,10),
        axis.text.x = element_text(vjust = -1),
        axis.title.x = element_text(vjust = -1))
}
```


```{r visualise trends over time}
# mean patch area
p_area_mn <- patch_plot(allpatch.ws$area_mn * 100 * 100, "Mean patch area (cm2)") + labs(x = "")
p_area_mn
# average patch area is stable for spor and sueda and increases over time for sarc 
# this makes sense, as spor is usually one large connected clump that grows from itself, sueda is generally lots of spaced out clumps, and sarc colonises an area by making lots of small clumps that expand into one another


# don't need mean class area, as I calculate this in the class area script


# mean core area - don't include as trends are exactly the same as patch area


# core area index (cai)
p_cai_mn <- patch_plot(allpatch.ws$cai_mn, "Mean core area index") + labs(x = "")
p_cai_mn
# core area index fluctuates for spor and sueda, possibly reflecting differences in classification?
# core area index increases steadily for sarc, probably reflecting increasing patch size
# this is the percentage of core area in relation to patch area


# number of patches
p_np <- patch_plot(allpatch.ws$np / 1000, "No. patches (x 1000)") + labs(x = "")
p_np
# variability probably reflects differences in classification
# however sarc is hump shaped as expected, in the middle lots of new separate clumps are forming, and they later merge
# spor and sueda are relatively consistent but variable
# sueda has more patches than spor even though its area is far smaller - probably because patch size is so small. However I think there should not be this many patches, sueda has been overestimated, and the small patch size is what made it difficult to pick up on these errors in validation. 

# cohesion
p_cohesion <- patch_plot(allpatch.ws$cohesion, "Cohesion (%)") + labs(x = "")
p_cohesion 
# connectedness of patches - locally aggregated (high values) or isolated (low values)
# high numbers for all - all are relatively aggregated
# spor is consistent and near 100% - as it grows it basically one interconnected clump
# sarc decreases and then increases - reflecting individual clump growth then connection
# sueda is consistently lower - reflecting that clumps do not generally touch one another. but increases over time - new clumps generally grow near existing ones

# euclidean nearest neighbour average distance between patches
p_enn <- patch_plot(allpatch.ws$enn_mn, "Mean ENN distance (m)") + labs(x = "")
p_enn
# distance to the nearest neighbouring patch
# greater distances for sueda - reflects that patches are more spread out
# sueda decreases over time (but variable) - new patches growing together
# sarc decreases somewhat and then increases - decrease reflects that patches are growing more closely together, increase might be because nearby patches merge and therefore have to find a further away patch to be measured against?
# spor is quite stable

# clumpiness
p_clumpy <- patch_plot(allpatch.ws$clumpy, "Clumpiness (0-1)") + labs(x = "")
p_clumpy 
# measure of how many adjacent patches are of the same class, relative to what would be expected if classes were randomly distributed
# sueda is lower - reflecting that patches are more spread out
# sueda and spor are consistent
# sarc initially decreases (new patches forming separate to one another), then increases (patches clumping together)

# contiguity
p_contig <- patch_plot(allpatch.ws$contig_mn, "Mean contiguity (0-1)")
p_contig
# shape metric reflecting connectivity of cells within patches
# spor and sueda are relatively consistent but variable over time, maybe reflecting classification differences
# sarc increases over time, suggesting patches become more regular shaped - perhaps strange shaped clumps merge together into more regular shaped clumps?

# mean shape index
p_shape_mn <- patch_plot(allpatch.ws$shape_mn, "Mean shape index")
p_shape_mn
# ratio between actual perimeter and theoretical minimum perimeter - ie roundness (1 = perfectly round, increases as becomes less round)
# sarc and spor are relatively consistent, decrease slightly over time - perhaps clumps initially grow in weird shapes and progress towards more circular shapes
# spor has greater values than sarc - sarc tends to grow in more regular shapes?
# sueda is around the same as sarc, more variable but then increases a lot at 46 mos - this may be from the classification rather than reflecting reality?


# landscape metrics

# contagions
p_contagion <- land_plot(allpatchl.w$contag, "Contagion (%)") 
p_contagion 
# measure of probability that adjacent cells belong to the same class, where 0 = cells unevenly distributed and 100 = cells perfectly randomly/evenly distributed
# related to how clumped together individual classes are and how interspersed different classes are
# decreases over time, potentially driven by sarc becoming more clumped together and more distinct from mud

# shannon diversity
p_shdi <- land_plot(allpatchl.w$shdi, "Shannon's diversity")
p_shdi
# related to the number of classes and abundance of each
# increases as the number of classes increases and as the evenness of classes increases
# over time, the landscape has become more diverse, potentially because the area of each class has become more similar due to sarc growth and mud decrease?

# shannon evenness
p_shei <- land_plot(allpatchl.w$shei, "Shannon's evenness (0-1)")
p_shei 
# ratio between actual shannon's diversity index and theoretical maximum shannon's diversity index- if all classes are evenly represented, would have a value of 1
# stable over time, increases at 46 months but I think this is probably a product of the classification rather than reality (ie there was more water at 46 mos)
```


```{r combine plots}
# combine class level patch plots with a common legend
ggarrange(p_area_mn, p_cai_mn, p_np, p_cohesion, p_enn, p_clumpy, p_contig, p_shape_mn, ncol = 2, nrow = 4, common.legend = TRUE, legend = "bottom", align = "hv") 
# export at 2100 x 2800

# combine landscape level plots
ggarrange(p_contagion, p_shdi, p_shei, ncol = 2, nrow = 2, align = "hv")
# export at 2100 x 1400
```


# Stats

Trends - Kendall rank correlations

Assumptions
- continous data - yes
- monotonic relationships - mostly, but when it's not, it's not going to be significant over time, so that's ok
```{r Kendall rank correlations}
allpatch.spor <- allpatch.ws %>% filter(category == "spor") %>% pad()
allpatch.sarc <- allpatch.ws %>% filter(category == "sarc") %>% pad()
allpatch.sueda <- allpatch.ws %>% filter(category == "sueda") %>% pad()
allpatchl.w <- allpatchl.w %>% pad()

# sporobolus
MannKendall(ts(allpatch.spor$area_mn,frequency = 365, start = c(2017,34)))
MannKendall(ts(allpatch.spor$cai_mn,frequency = 365, start = c(2017,34)))
MannKendall(ts(allpatch.spor$np,frequency = 365, start = c(2017,34)))
MannKendall(ts(allpatch.spor$cohesion,frequency = 365, start = c(2017,34)))
MannKendall(ts(allpatch.spor$enn_mn,frequency = 365, start = c(2017,34)))
MannKendall(ts(allpatch.spor$clumpy,frequency = 365, start = c(2017,34)))
MannKendall(ts(allpatch.spor$contig_mn,frequency = 365, start = c(2017,34)))
MannKendall(ts(allpatch.spor$shape_mn,frequency = 365, start = c(2017,34)))
# all non-significant

# sarcorcornia
MannKendall(ts(allpatch.sarc$area_mn,frequency = 365, start = c(2017,34))) # sig
MannKendall(ts(allpatch.sarc$cai_mn,frequency = 365, start = c(2017,34))) # sig
MannKendall(ts(allpatch.sarc$np,frequency = 365, start = c(2017,34)))
MannKendall(ts(allpatch.sarc$cohesion,frequency = 365, start = c(2017,34))) # sig
MannKendall(ts(allpatch.sarc$enn_mn,frequency = 365, start = c(2017,34)))
MannKendall(ts(allpatch.sarc$clumpy,frequency = 365, start = c(2017,34)))
MannKendall(ts(allpatch.sarc$contig_mn,frequency = 365, start = c(2017,34))) # sig
MannKendall(ts(allpatch.sarc$shape_mn,frequency = 365, start = c(2017,34)))

# suaeda
MannKendall(ts(allpatch.sueda$area_mn,frequency = 365, start = c(2018,313)))
MannKendall(ts(allpatch.sueda$cai_mn,frequency = 365, start = c(2018,313)))
MannKendall(ts(allpatch.sueda$np,frequency = 365, start = c(2018,313)))
MannKendall(ts(allpatch.sueda$cohesion,frequency = 365, start = c(2018,313)))
MannKendall(ts(allpatch.sueda$enn_mn,frequency = 365, start = c(2018,313)))
MannKendall(ts(allpatch.sueda$clumpy,frequency = 365, start = c(2018,313)))
MannKendall(ts(allpatch.sueda$contig_mn,frequency = 365, start = c(2018,313)))
MannKendall(ts(allpatch.sueda$shape_mn,frequency = 365, start = c(2018,313)))

# landscape
MannKendall(ts(allpatchl.w$contag,frequency = 365, start = c(2017,34)))
MannKendall(ts(allpatchl.w$shdi,frequency = 365, start = c(2017,34))) # sig 
MannKendall(ts(allpatchl.w$shei,frequency = 365, start = c(2017,34)))
```


Differences between groups - Repeated measures ANOVAs

Assumptions of repeated measures ANOVAs
- independence- observations are independent (taken at diff times)
- normality
- sphericity

```{r}

```

